---
title: "tidy-model"
author: "betty yinges"
date: "2025-02-27"
output: html_document
---

```{r setup, include=FALSE}
## Setup & Loading Packages
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(modeldata)
library(ggplot2)

```

## Introduction

`This R Markdown document demonstrates my abilities to use models for data analysis using a data set collected on crickets.`

```{r}
data(crickets, package = "modeldata")
names(crickets)
```

## What is that sound?
`In this report, I examine what factors predict a cricket's chirp rate.`

```{r}
summary(crickets)

ggplot(crickets, aes(x = rate)) + 
  geom_histogram(bins = 10, fill = "skyblue", color = "black") + 
  ggtitle("Distribution of Chirp Rates") + 
  xlab('Chirp Rate (per min.)') + 
  theme_minimal()

```
```
Observations:

- The chirp rate varies among crickets.
- The distribution helps us understand the variation in chirping frequency.
```

## Temperature affects chirp rate
```
I now investigate the relationship between temperature and chirp rate.

```
```{r}
ggplot(crickets, aes(x = temp, y = rate)) + 
  geom_jitter(alpha = 0.6, color = "blue") + 
  geom_smooth(method = 'lm', color = "red", fill = "lightgray") + 
  ggtitle("Temperature vs. Chirp Rate") + 
  ylab('Chirp Rate (per min.)') + 
  xlab('Temperature (°C)') + 
  theme_minimal()

# Fit linear model: chirp rate ~ temperature
temp_lm <- lm(rate ~ temp, data = crickets)

# Display model summary
summary(temp_lm)


```
`Findings:`

`Chirp rate increases as temperature increases.`

`Based on a scatter plot of temperature and chirping and a correlation test, it seems that as temperature increases one degree, the rate of chirping increases about 4.2 chirps per minute.` 

## Species-specific effects of temperature on chirping

`Now, I check if species differences affect chirp rate.`

## Scatter Plot with Species Coloring
```{r}
ggplot(crickets, aes(x = temp, y = rate, color = species)) + 
  geom_jitter(alpha = 0.6) + 
  geom_smooth(method = 'lm', se = FALSE) + 
  ggtitle("Temperature vs. Chirp Rate by Species") + 
  ylab('Chirp Rate (per min.)') + 
  xlab('Temperature (°C)') + 
  theme_minimal()

# Fit linear model including species
species_lm <- lm(rate ~ temp + species, data = crickets)

# Display model summary
summary(species_lm)
```

`Key Insights:`

`Different species have different chirp rates.`
`The model shows that O. exclamationis chirps more frequently than O. niveus.`




## Exploring Interactions Between Temperature & Species
`Here, I investigate whether temperature effects vary between species.`

## Chirp Rate Histogram by Species

```{r}
ggplot(crickets, aes(x = rate, fill = species)) + 
  geom_histogram(position = 'identity', alpha = 0.7, bins = 10) + 
  ggtitle("Chirp Rate Distribution by Species") + 
  xlab('Chirp Rate (per min.)') + 
  theme_minimal()

```
## Interaction Model: Temperature & Species
```{r}
ggplot(crickets, aes(x = temp, y = rate, color = species)) + 
  geom_jitter(alpha = 0.6) + 
  geom_smooth(method = 'lm', se = FALSE) + 
  ggtitle("Temperature vs. Chirp Rate (with Interaction)") + 
  ylab('Chirp Rate (per min.)') + 
  xlab('Temperature (°C)') + 
  theme_minimal()

```
`Fitting an Interaction Model`
```{r}
# Fit a model with interaction term
species_x_temp_lm <- lm(rate ~ temp * species, data = crickets)

# Display model summary
summary(species_x_temp_lm)

# Compare with previous species model using ANOVA
anova(species_lm, species_x_temp_lm)
```
`Findings:`

`The interaction model does not significantly improve predictions.`
`The simpler species_lm model is sufficient.`


## Diamond Price Analysis
```
`Using these skills to a different dataset: the diamonds dataset    

The following model predicts diamond prices based on carat weight, cut, clarity, and color.
```

```{r}
data(diamonds, package = "ggplot2")

ggplot(diamonds, aes(x = carat, y = price)) + 
  geom_point(alpha = 0.3, color = "purple") + 
  geom_smooth(method = 'lm', se = FALSE, color = "red") + 
  ggtitle("Diamond Carat vs. Price") + 
  xlab('Carat') + 
  ylab('Price ($)') + 
  theme_minimal()
diamonds_lm <- lm(price ~ carat + cut + clarity + color, data = diamonds)
summary(diamonds_lm)

```
```
The linear model suggests that carat weight is the strongest predictor of diamond price,
while cut, clarity, and color also contribute to variations in price.
```
```{r}
# Fit a linear model predicting diamond prices
diamonds_lm <- lm(price ~ carat + cut + clarity + color, data = diamonds)

# Display model summary
summary(diamonds_lm)


```

```
Key Takeaways:

Carat weight is the strongest predictor of diamond price.
Cut, clarity, and color also impact pricing. 
```
## Conclusion

```
This analysis demonstrates how linear regression models can be used to study relationships in data, such as:
✅ Temperature affecting chirp rate in crickets.
✅ Species differences in chirp rate.
✅ Predicting diamond prices based on multiple factors.
```
